---
title: "Test harness for `go_enrichment.R`"
subtitle: "Unit tests, regression tests for known bugs, and end-to-end CLI runs"
author: "Brannan Lab"
date: "`r format(Sys.time(), '%Y-%m-%d %H:%M')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    df_print: paged
---

```{r knitr-setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  collapse = TRUE, comment = "#>"
)
```

> **If `rmarkdown::render()` complains about pandoc**, there is no pandoc on the
> login node. Either run the chunks interactively in VS Code / RStudio (the
> intended use), or produce a plain markdown report with
> `knitr::knit("test_go_enrichment.Rmd")`.

# What this document is for

`go_enrichment.R` has never been executed — the README says so explicitly
(README:233). This harness does three separate jobs, and you can run them
independently:

| Part | Needs Bioconductor? | Purpose |
|---|---|---|
| **1. Static + unit tests** | no | Parse the script, pull its helper functions out, test them in isolation |
| **2. Regression tests** | no | Reproduce each confirmed bug so you can prove a fix worked |
| **3. End-to-end CLI runs** | **yes** | Actually invoke the script on realistic fixtures and validate the outputs |

Every suggested change is flagged inline as **`EDIT #n`** (a defect) or
**`IMPROVEMENT #n`** (a quality/robustness point), with the source line number.
They are collected into a single table in the [Summary of suggested edits] section,
and section [Ready-to-apply patches] holds copy-pasteable replacement code.

---

# 0. Configuration

```{r config}
SCRIPT   <- "/condo/brannanlab/tmhaxs421/scripts/ScriptsToolkit/analyses/gene_ontology/go_enrichment.R"

# A real edgeR topTags export from the RPS2-STAMP drug study. Used both as a
# realistic fixture and as the trigger for EDIT #1.
REAL_EDGER <- paste0("/condo/brannanlab/tmhaxs421/Bulk_RiboStamp/ARVSDrugStudy/",
                     "results/08_edgeR/MCF10A/TX_drug_pooled.csv")

# Everything this document writes goes here. Safe to delete afterwards.
TESTROOT <- file.path(dirname(SCRIPT), "test_run")
FIXDIR   <- file.path(TESTROOT, "fixtures")
OUTDIR   <- file.path(TESTROOT, "outputs")
dir.create(FIXDIR, recursive = TRUE, showWarnings = FALSE)
dir.create(OUTDIR, recursive = TRUE, showWarnings = FALSE)

ORGANISM <- "human"   # fixtures below use human gene symbols
stopifnot(file.exists(SCRIPT))
cat("script:  ", SCRIPT, "\n")
cat("test dir:", TESTROOT, "\n")
```

## Minimal test harness

Deliberately dependency-free so Part 1 and Part 2 run anywhere.

```{r harness}
.results <- new.env(parent = emptyenv())
.results$rows <- list()

.record <- function(part, label, status) {
  .results$rows[[length(.results$rows) + 1L]] <- data.frame(
    part = part, test = label, status = status, stringsAsFactors = FALSE)
  cat(sprintf("[%s] %s\n", status, label))
  invisible(identical(status, "PASS"))
}

chk <- function(label, expr, expect = TRUE, part = "") {
  got  <- tryCatch(isTRUE(expr), error = function(e) NA)
  pass <- !is.na(got) && identical(got, isTRUE(expect))
  .record(part, label, if (pass) "PASS" else "FAIL")
}

skip <- function(label, why, part = "") .record(part, paste0(label, " [", why, "]"), "SKIP")

test_report <- function() {
  if (!length(.results$rows)) return(invisible(NULL))
  df <- do.call(rbind, .results$rows)
  cat(sprintf("\n%d tests: %d PASS, %d FAIL, %d SKIP\n", nrow(df),
              sum(df$status == "PASS"), sum(df$status == "FAIL"), sum(df$status == "SKIP")))
  knitr::kable(df, row.names = FALSE)
}
```

## Environment audit

```{r env-audit}
pkgs <- c("optparse", "clusterProfiler", "GO.db", "ggplot2", "topGO",
          "GOSemSim", "DOSE", "org.Hs.eg.db")
have <- vapply(pkgs, function(p) requireNamespace(p, quietly = TRUE), logical(1))
env_tbl <- data.frame(package = pkgs,
                      version = vapply(pkgs, function(p)
                        tryCatch(as.character(packageVersion(p)),
                                 error = function(e) NA_character_), character(1)),
                      installed = have, row.names = NULL)
print(env_tbl)

HAVE_BIOC <- all(have[c("optparse", "clusterProfiler", "GO.db", "ggplot2", "org.Hs.eg.db")])
HAVE_TOPGO <- have[["topGO"]]
cat("\nR:", R.version.string, "\n")
cat("End-to-end CLI tests will", if (HAVE_BIOC) "RUN" else "be SKIPPED (install the stack first)", "\n")
```

> **IMPROVEMENT #1 — package the environment.**
> Every other tool in `ScriptsToolkit/` has a conda recipe in `environments/`.
> This one does not, which is why the script has never run. Add
> `environments/gene_ontology.yaml`:
>
> ```yaml
> name: gene_ontology
> channels: [conda-forge, bioconda]
> dependencies:
>   - r-base>=4.3
>   - r-optparse
>   - r-ggplot2
>   - bioconductor-clusterprofiler
>   - bioconductor-go.db
>   - bioconductor-gosemsim
>   - bioconductor-topgo
>   - bioconductor-org.hs.eg.db
>   - bioconductor-org.mm.eg.db
> ```
>
> **IMPROVEMENT #2 — `GOSemSim` is not an optional dependency.**
> README:45-46 says "`topGO` and `GOSemSim` are only needed for `--topgo` and
> `--simplify`". `GOSemSim` is an `Imports:` of clusterProfiler, so it is always
> installed. Harmless, but the sentence should name only `topGO`. Conversely,
> [go_enrichment.R:150-151] adds `topGO` to `need` but the README install block
> lists it as unconditional — the two disagree in opposite directions.

---

# 1. Static analysis and unit tests

## 1.1 The script parses

```{r parse}
parsed <- tryCatch(parse(SCRIPT), error = function(e) e)
chk("script parses without syntax error", !inherits(parsed, "error"), part = "static")
cat("top-level expressions:", length(parsed), "\n")
```

## 1.2 Extract the helper functions from source

Rather than copying the helpers into this document (which would drift), we
evaluate *only* the top-level `name <- function(...)` definitions from the real
script into a sandbox, and stub the globals they close over. Tests below
therefore always exercise the current source.

```{r extract-helpers}
sandbox <- new.env(parent = globalenv())

# Stubs for the globals the helpers reference.
sandbox$fail <- function(...) stop(sprintf(...), call. = FALSE)
sandbox$msg  <- function(...) cat(sprintf(...), "\n", sep = "")
sandbox$.opts <- list(ont = "BP", `min-size` = 10, `max-size` = 500,
                      `p-method` = "BH", padj = 0.05, `top-n` = 25,
                      kappa = 0.3, `heatmap-top` = 20, prefix = "TEST",
                      outdir = OUTDIR)
sandbox$g <- function(k) sandbox$.opts[[k]]

is_fundef <- function(e) is.call(e) && identical(as.character(e[[1]]), "<-") &&
  is.call(e[[3]]) && identical(as.character(e[[3]][[1]]), "function")

fundefs <- Filter(is_fundef, as.list(parsed))
for (e in fundefs) eval(e, envir = sandbox)
extracted <- vapply(fundefs, function(e) as.character(e[[2]]), character(1))
cat("extracted", length(extracted), "helper functions:\n ",
    paste(extracted, collapse = ", "), "\n")
chk("core helpers were extracted",
    all(c("read_gene_file", "read_table_auto", "guess_id_type", "kappa_sim",
          "go_depth", "ratio_num", "ratio_den") %in% extracted), part = "static")
```

## 1.3 `ratio_num` / `ratio_den` and fold enrichment

```{r test-ratios}
chk("ratio_num('12/300') == 12", sandbox$ratio_num("12/300") == 12, part = "unit")
chk("ratio_den('12/300') == 300", sandbox$ratio_den("12/300") == 300, part = "unit")
chk("ratio_* are vectorised",
    all(sandbox$ratio_num(c("3/10", "7/20")) == c(3, 7)), part = "unit")

# Fold enrichment as annotate_terms() computes it, checked against first principles
k <- 12; n <- 300; M <- 40; N <- 12000
fe_script <- (k / n) / (M / N)
chk("FoldEnrichment == (k/n)/(M/N)", isTRUE(all.equal(fe_script, (12/300)/(40/12000))),
    part = "unit")
cat("fold enrichment:", round(fe_script, 3), "\n")
```

## 1.4 `read_gene_file`

```{r test-read-gene-file}
f1 <- file.path(FIXDIR, "plain_list.txt")
writeLines(c("TP53", "BRCA1", "", "  EGFR  ", "MYC", "TP53"), f1)
r1 <- sandbox$read_gene_file(f1)
chk("blank lines dropped and whitespace trimmed", identical(r1, c("TP53","BRCA1","EGFR","MYC")),
    part = "unit")

f2 <- file.path(FIXDIR, "with_header.csv")
writeLines(c("gene_id,logFC", "\"TP53\",1.2", "\"BRCA1\",-0.4"), f2)
chk("header line dropped, first field taken, quotes stripped",
    identical(sandbox$read_gene_file(f2), c("TP53", "BRCA1")), part = "unit")

f3 <- file.path(FIXDIR, "tsv_list.tsv")
writeLines(c("symbol\tpadj", "TP53\t0.01", "MYC\t0.02"), f3)
chk("tab-separated table handled",
    identical(sandbox$read_gene_file(f3), c("TP53", "MYC")), part = "unit")
```

> **IMPROVEMENT #3 — the header heuristic is narrow.**
> [go_enrichment.R:174] only recognises `gene|gene_id|gene_name|symbol|id` as a
> header. A file starting with `Gene.ID`, `GeneSymbol`, `ensembl_gene_id`, or an
> empty first cell keeps its header as a gene, which then silently fails to map.
> Since the mapping rate is printed, one lost gene out of 2000 is invisible.

```{r test-read-gene-file-improve}
f4 <- file.path(FIXDIR, "odd_header.csv")
writeLines(c("ensembl_gene_id,padj", "ENSG00000141510,0.01", "ENSG00000012048,0.02"), f4)
r4 <- sandbox$read_gene_file(f4)
chk("IMPROVEMENT #3: 'ensembl_gene_id' header is NOT stripped (documents current behaviour)",
    "ensembl_gene_id" %in% r4, part = "unit")
print(r4)
```

## 1.5 `kappa_sim` — verified correct

This is the part of the script most likely to hide a silent index error, so it is
checked against an independent 2x2 implementation.

```{r test-kappa}
ref_kappa <- function(g1, g2, universe) {
  x <- universe %in% g1; y <- universe %in% g2
  a <- sum(x & y); b <- sum(x & !y); c <- sum(!x & y); d <- sum(!x & !y)
  n <- length(universe)
  po <- (a + d) / n
  pe <- ((a + b) * (a + c) + (c + d) * (b + d)) / n^2
  (po - pe) / (1 - pe)
}

set.seed(7)
U  <- paste0("G", 1:60)
tg <- list(T1 = sample(U, 20), T2 = sample(U, 25), T3 = sample(U, 15),
           T4 = NULL)
tg$T4 <- tg$T1[1:18]                      # near-identical to T1 -> high kappa
K <- sandbox$kappa_sim(tg, U)

pairs <- combn(names(tg), 2, simplify = FALSE)
agree <- vapply(pairs, function(p)
  isTRUE(all.equal(K[p[1], p[2]], ref_kappa(tg[[p[1]]], tg[[p[2]]], U))), logical(1))
chk("kappa matches an independent 2x2 implementation for all pairs", all(agree), part = "unit")
chk("kappa matrix is symmetric", isTRUE(all.equal(K, t(K))), part = "unit")
chk("diagonal is 1", all(diag(K) == 1), part = "unit")
chk("near-identical terms give high kappa", K["T1", "T4"] > 0.85, part = "unit")
round(K, 3)
```

> **The kappa implementation is correct.** The `Ri`/`Rj` marginal construction at
> [go_enrichment.R:380-384] is right, including the transpose. No change needed.

> **IMPROVEMENT #4 — the kappa universe deviates from Metascape.**
> [go_enrichment.R:406] sets `universe` to the union of genes appearing in
> *significant terms*. Metascape computes kappa over the **input gene list**.
> That changes cell `d` of every 2x2 table and therefore every kappa value, so
> the `--kappa 0.3` cutoff inherited from Metascape is not directly transferable.
> The README is candid about the tree-cut approximation (README:108-110); this
> belongs in the same paragraph. Demonstration:

```{r kappa-universe-effect}
small_U <- unique(unlist(tg))            # what the script uses
big_U   <- U                             # what Metascape would use
cat("kappa(T1,T2) with term-union universe (n=", length(small_U), "): ",
    round(ref_kappa(tg$T1, tg$T2, small_U), 4), "\n", sep = "")
cat("kappa(T1,T2) with full-list universe (n=", length(big_U), "): ",
    round(ref_kappa(tg$T1, tg$T2, big_U), 4), "\n", sep = "")
```

## 1.6 `go_depth`

```{r test-go-depth, eval=HAVE_BIOC}
d <- sandbox$go_depth(c("GO:0008150",   # biological_process (root)
                        "GO:0006915",   # apoptotic process
                        "GO:0097190"),  # apoptotic signaling pathway
                      "BP")
names(d) <- c("root", "apoptotic process", "apoptotic signaling pathway")
print(d)
chk("BP root has depth 0", d[[1]] == 0, part = "unit")
chk("depth increases down the ontology", d[[2]] > 0 && d[[3]] > d[[2]], part = "unit")
chk("unknown GO ID yields NA", is.na(sandbox$go_depth("GO:9999999", "BP")), part = "unit")
chk("ont='ALL' returns NA (documented limitation)",
    all(is.na(sandbox$go_depth(c("GO:0006915"), "ALL"))), part = "unit")
```

```{r go-depth-skip, eval=!HAVE_BIOC, echo=FALSE}
cat("SKIPPED: GO.db not installed\n")
```

---

# 2. Regression tests for confirmed defects

Each of these **currently fails** (or silently misbehaves). After applying the
patches in [Ready-to-apply patches], re-run this section — they should all pass.

## 2.1 EDIT #1 (critical) — edgeR exports with gene IDs in rownames are silently emptied

This is the one that hits your real data. `topTags()` tables written with
`write.csv(..., row.names = TRUE)` have an **empty first header field**:

```{r edit1-show}
cat(readLines(REAL_EDGER, n = 2), sep = "\n")
```

`read_table_auto` with `check.names = FALSE` keeps that column named `""`.
Then at [go_enrichment.R:252] `gcol <- names(de)[1]` resolves to `""`; the
validation at [go_enrichment.R:253-255] **passes**, because `"" %in% names(de)`
is `TRUE`; and `de[[""]]` returns `NULL`.

```{r edit1-repro}
de <- sandbox$read_table_auto(REAL_EDGER)
gcol <- names(de)[1]

cat("ncol:", ncol(de), "\n")
cat("column names:", paste(sprintf("[%s]", names(de)), collapse = " "), "\n")
cat("gcol resolves to: [", gcol, "]\n", sep = "")
cat("validation `gcol %in% names(de)`:", gcol %in% names(de), " <- passes!\n")
cat("class(de[[gcol]]):", class(de[[gcol]]), "\n")

# now replay lines 256-259 of the script
de_f <- de[!is.na(de[[gcol]]) & nzchar(de[[gcol]]), , drop = FALSE]
de_f <- de_f[!duplicated(de_f[[gcol]]), , drop = FALSE]
cat("\nrows in file:", nrow(de), "  rows surviving the script's filter:", nrow(de_f), "\n")

chk("EDIT #1: real edgeR export retains its rows",
    nrow(de_f) > 1000, part = "regression")
```

**Impact.** The whole table collapses to zero rows *before* anything is
reported, so `bg_raw` is empty and the run dies later inside `to_entrez` with a
message about ID mapping — pointing at the wrong problem entirely. It never
warns that it dropped 15,179 genes.

The README's workaround (README:156-163) tells you to add a `gene_id` column
first, which does avoid it — but nothing in the script *enforces* that, and the
failure mode is silent and misleading.

> **EDIT #1.** In `read_table_auto`, promote an empty/`X`/`X.1` first column name
> to a real one, or fold `row.names` in. And in the validation block, reject an
> empty `gcol` and check `!is.null(de[[gcol]])` rather than name membership.
> Patch in [Ready-to-apply patches].

```{r edit1-proposed}
read_table_auto_fixed <- function(path) {
  if (!file.exists(path)) stop("file not found: ", path)
  sep <- if (grepl("\\.csv$", path, ignore.case = TRUE)) "," else "\t"
  df  <- read.delim(path, sep = sep, header = TRUE, check.names = FALSE,
                    stringsAsFactors = FALSE)
  if (ncol(df) == 1)
    df <- read.delim(path, sep = ",", header = TRUE, check.names = FALSE,
                     stringsAsFactors = FALSE)
  # edgeR/DESeq2 exports written with row.names=TRUE have a blank first header
  if (!nzchar(trimws(names(df)[1])) || names(df)[1] %in% c("X", "V1", "...1"))
    names(df)[1] <- "gene_id"
  df
}
de2 <- read_table_auto_fixed(REAL_EDGER)
cat("first column now named:", names(de2)[1], "\n")
cat("rows retained:", nrow(de2[!is.na(de2[[1]]) & nzchar(de2[[1]]), , drop = FALSE]), "\n")
chk("EDIT #1 fix: rows retained with patched reader",
    nrow(de2[nzchar(de2[[1]]), , drop = FALSE]) > 1000, part = "regression")
```

## 2.2 EDIT #2 — `--topgo` with `--allow-genome-background` crashes

`bg_entrez` is `NULL` when no background is supplied
([go_enrichment.R:302](#)). `enrichGO` tolerates `universe = NULL`; `run_topgo`
does not, because `NULL %in% x` is `logical(0)`.

```{r edit2-repro}
bg_entrez <- NULL
gl <- factor(as.integer(bg_entrez %in% c("7157", "672")), levels = c(0, 1))
cat("length of topGO allGenes factor:", length(gl), "\n")
chk("EDIT #2: allGenes factor is non-empty when background is NULL",
    length(gl) > 0, part = "regression")
```

> **EDIT #2.** In the ORA loop, build a fallback universe before calling
> `run_topgo`: `bg_topgo <- if (is.null(bg_entrez)) keys(OrgDb, "ENTREZID") else bg_entrez`.

## 2.3 EDIT #3 — ID auto-detection breaks when nothing is significant

[go_enrichment.R:285-288] guesses the ID type from `unlist(sets)` — the
*significant* genes only. A DE table with no hits gives an empty vector, which
falls through to `"SYMBOL"`. `--gsea` then runs anyway and calls
`bitr(fromType = "SYMBOL")` on what may be Ensembl IDs.

```{r edit3-repro}
sets_empty <- list(up = character(0), down = character(0))
guessed <- sandbox$guess_id_type(unlist(sets_empty, use.names = FALSE))
cat("guess_id_type(<no significant genes>) ->", guessed, "\n")
chk("EDIT #3: ID type is not silently mis-guessed on an empty query",
    !identical(guessed, "SYMBOL"), part = "regression")

# the same table's background would have given the right answer
cat("guess from background instead ->",
    sandbox$guess_id_type(c("ENSG00000141510", "ENSG00000012048", "ENSG00000139618")), "\n")
```

This is exactly the scenario the README flags via Stead et al. (README:191-193):
an empty DE result plus a full GSEA hit list is a warning sign — but here the
GSEA would not even reach the right genes.

> **EDIT #3.** Guess from the background when one exists, falling back to `sets`:
> `id_type <- guess_id_type(if (!is.null(bg_raw)) bg_raw else unlist(sets, use.names = FALSE))`

## 2.4 EDIT #4 — `guess_id_type` indexes past the filtered vector

[go_enrichment.R:192]: `x <- x[!is.na(x)][1:min(200, length(x))]`. The RHS
`length(x)` is the **pre-filter** length, so `NA`s are re-introduced and dilute
the `mean(grepl(...))` fractions.

```{r edit4-repro}
ens <- sprintf("ENSG%011d", 1:100)
chk("EDIT #4: 100 Ensembl IDs + 101 NAs still detected as ENSEMBL",
    sandbox$guess_id_type(c(ens, rep(NA_character_, 101))) == "ENSEMBL", part = "regression")
chk("EDIT #4: clean input is detected correctly (control)",
    sandbox$guess_id_type(ens) == "ENSEMBL", part = "regression")

# and the empty-vector edge
cat("1:min(200, 0) ->", paste(1:min(200, 0), collapse = ", "), "  <- indexes element 1 of nothing\n")
```

Currently latent, because both input paths strip `NA`s upstream — but one
refactor away from mattering.

> **EDIT #4.** `x <- x[!is.na(x) & nzchar(x)]; x <- head(x, 200)`

## 2.5 EDIT #5 — `--rank-metric stat` accepts an unsigned column silently

`is_unsigned` is computed at [go_enrichment.R:554] but only consulted inside the
`auto` branch ([go_enrichment.R:562-567]). So `--rank-metric stat --stat-col F`
ranks on unsigned F, collapsing up- and down-regulated genes into the same end of
the list — the precise error the README's edgeR section exists to prevent.

```{r edit5-repro}
unsigned_cols <- c("F", "LR", "f", "lr", "Chisq")
statcol <- "F"; rmet <- "stat"
is_unsigned <- statcol %in% unsigned_cols
cat("statcol =", statcol, " is_unsigned =", is_unsigned,
    " rmet =", rmet, " -> metric used as-is, sign discarded\n")
chk("EDIT #5: unsigned stat column with --rank-metric stat is caught",
    !(rmet == "stat" && is_unsigned), part = "regression")

# what it does to the ranking
set.seed(1)
lfc <- c(3, -3, 1.5, -1.5); Fstat <- c(90, 88, 20, 19)
cat("\nwrong (unsigned F):    ", paste(round(Fstat, 1), collapse = ", "), "\n")
cat("right (sign(lfc)*sqrt):", paste(round(sign(lfc) * sqrt(Fstat), 2), collapse = ", "), "\n")
```

> **EDIT #5.** Promote to `signed_sqrt` with a warning when the user asks for
> `stat` on a column in `unsigned_cols`.

## 2.6 EDIT #6 — `NA` in `p.adjust` injects an all-`NA` row

[go_enrichment.R:494-495] subsets with a raw logical, so an `NA` propagates into
a literal `NA` row in `*_significant.tsv`.

```{r edit6-repro}
res  <- data.frame(ID = c("GO:1","GO:2","GO:3"), p.adjust = c(0.01, NA, 0.2), Count = c(5,5,5))
keep <- res$p.adjust < 0.05 & res$Count >= 3
sig  <- res[keep, , drop = FALSE]
print(sig)
chk("EDIT #6: no NA rows in the significant table",
    !any(is.na(sig$ID)), part = "regression")

sig_fixed <- res[which(keep), , drop = FALSE]   # the fix
chk("EDIT #6 fix: which() removes the NA row", !any(is.na(sig_fixed$ID)), part = "regression")
```

## 2.7 EDIT #7 — the ranked list reaches `gseGO` as a 1-d array

`tapply` returns an array and `sort()` preserves `dim`.

```{r edit7-repro}
metric <- c(3, -5, 2, 7); ent <- c("1", "1", "2", "3")
rk <- sort(tapply(metric, ent, function(v) v[which.max(abs(v))]), decreasing = TRUE)
cat("class:", class(rk), " dim:", paste(dim(rk), collapse = ","), "\n")
chk("EDIT #7: ranked vector has no dim attribute", is.null(dim(rk)), part = "regression")

rk_fixed <- setNames(as.numeric(rk), names(rk))   # the fix
chk("EDIT #7 fix: plain named numeric", is.null(dim(rk_fixed)) && is.numeric(rk_fixed),
    part = "regression")
```

Low risk — `fgsea` almost certainly copes — but the fix is one line and free.

## 2.8 EDIT #8 — duplicate `Description` values would error

[go_enrichment.R:351] and [:442] build factors with possibly-duplicated `levels`.
Reachable under `--ont ALL`, where a term name can repeat across domains.

```{r edit8-repro}
r <- tryCatch(factor(c("a", "a"), levels = rev(c("a", "a"))),
              error = function(e) conditionMessage(e))
cat("factor() with duplicate levels ->", if (is.character(r)) paste("ERROR:", r) else "ok", "\n")
chk("EDIT #8: duplicate descriptions do not error", !is.character(r), part = "regression")
```

> **EDIT #8.** Use `unique()` on the level vector, or disambiguate with the GO ID:
> `levels = rev(unique(df$Description[order(df$FoldEnrichment)]))`

## 2.9 EDIT #9 — `is_representative` mislabels rows

[go_enrichment.R:424-426] truncates `reps` to `--heatmap-top` **before** setting
the flag, so clusters outside the top N have no representative marked at all.
The column currently means "shown in the heatmap", not "cluster representative".

```{r edit9-repro}
set.seed(3)
memb <- data.frame(cluster = rep(1:5, each = 3), ID = paste0("GO:", 1:15),
                   best_padj = runif(15))
memb <- memb[order(memb$cluster, memb$best_padj), ]
reps <- memb[!duplicated(memb$cluster), ]
reps <- head(reps[order(reps$best_padj), ], 2)          # --heatmap-top 2
memb$is_representative <- memb$ID %in% reps$ID
cat("clusters:", length(unique(memb$cluster)),
    "  rows flagged representative:", sum(memb$is_representative), "\n")
chk("EDIT #9: every cluster has exactly one representative",
    sum(memb$is_representative) == length(unique(memb$cluster)), part = "regression")
```

## 2.10 EDIT #10 — the mapping-rate line is internally inconsistent

[go_enrichment.R:215-217] prints `length(out)` (unique **Entrez**) as the count
but computes the percentage from unique **source** IDs. Under many-to-one or
one-to-many mapping the printed `X -> Y (Z%)` has `Z != Y/X`. Since the README
frames this as a reproducibility feature (README:77-78), it should be consistent.

```{r edit10-repro}
# a symbol mapping where two symbols collapse onto one Entrez ID
m <- data.frame(SYMBOL = c("A", "B", "C", "D"), ENTREZID = c("1", "1", "2", "3"))
ids <- c("A", "B", "C", "D", "E")            # E fails to map
out <- unique(m$ENTREZID)
cat(sprintf("printed: %d input -> %d mapped (%.1f%%)\n",
            length(ids), length(out), 100 * length(unique(m$SYMBOL)) / length(ids)))
cat("         3 != 80%% of 5 -- the count and the percentage describe different things\n")
chk("EDIT #10: reported count and percentage agree",
    abs(100 * length(out) / length(ids) - 100 * length(unique(m$SYMBOL)) / length(ids)) < 1e-9,
    part = "regression")
```

---

# 3. Fixtures

Built from the real MCF10A edgeR table so gene symbols are genuine and mapping
rates are realistic.

```{r fixtures}
raw <- read.csv(REAL_EDGER, check.names = FALSE, stringsAsFactors = FALSE)
names(raw)[1] <- "gene_id"
raw <- raw[!is.na(raw$gene_id) & nzchar(raw$gene_id), ]
cat("source table:", nrow(raw), "genes,", ncol(raw), "columns\n")

## (a) edgeR table, correctly exported (gene IDs in a real column)
edger_fix <- file.path(FIXDIR, "edger_fixed.csv")
write.csv(raw, edger_fix, row.names = FALSE)

## (b) edgeR table exactly as topTags writes it (reproduces EDIT #1)
edger_raw <- file.path(FIXDIR, "edger_rownames.csv")
file.copy(REAL_EDGER, edger_raw, overwrite = TRUE)

## (c) DESeq2-shaped table: recover the signed moderated statistic from F
deseq_like <- data.frame(
  gene_id        = raw$gene_id,
  baseMean       = 2^raw$logCPM * 10,
  log2FoldChange = raw$logFC,
  lfcSE          = abs(raw$logFC) / pmax(sqrt(pmax(raw$F, 0)), 1e-6),
  stat           = sign(raw$logFC) * sqrt(pmax(raw$F, 0)),
  pvalue         = raw$PValue,
  padj           = raw$FDR,
  stringsAsFactors = FALSE)
deseq_csv <- file.path(FIXDIR, "deseq2_like.csv")
write.csv(deseq_like, deseq_csv, row.names = FALSE)

## (d) plain gene lists + background (the CLIP / peak-gene case)
sig  <- !is.na(raw$FDR) & raw$FDR < 0.05
up   <- raw$gene_id[sig & raw$logFC >=  1]
down <- raw$gene_id[sig & raw$logFC <= -1]
bg   <- raw$gene_id
writeLines(head(up,   400), file.path(FIXDIR, "up.txt"))
writeLines(head(down, 400), file.path(FIXDIR, "down.txt"))
writeLines(bg,              file.path(FIXDIR, "background.txt"))

## (e) four overlapping lists for the Metascape-style heatmap
set.seed(42)
pool <- head(up, 600)
for (i in 1:4) {
  li <- unique(c(sample(pool, 180), sample(bg, 60)))
  writeLines(li, file.path(FIXDIR, sprintf("list%d.txt", i)))
}

## (f) a DE table with ZERO significant genes (drives EDIT #3 end-to-end)
null_de <- deseq_like; null_de$padj <- 1; null_de$pvalue <- 1
null_csv <- file.path(FIXDIR, "deseq2_null.csv")
write.csv(null_de, null_csv, row.names = FALSE)

cat("\nfixtures written to", FIXDIR, "\n")
cat("  up:", length(up), " down:", length(down), " background:", length(bg), "\n")
print(data.frame(file = list.files(FIXDIR), row.names = NULL))
```

> **IMPROVEMENT #5 — undocumented gene-selection defaults.**
> `--lfc-cutoff` (default **1**, i.e. 2-fold) and `--padj-cutoff` (0.05) define
> what "up" and "down" *mean*, but neither appears in the README's
> "Defaults and when to change them" table (README:218-230). That table is where
> a reader looks to understand the analysis. Note below how much the gene sets
> move:

```{r lfc-sensitivity}
sens <- do.call(rbind, lapply(c(0, 0.5, 1, 1.5), function(cut) {
  s <- !is.na(raw$FDR) & raw$FDR < 0.05 & abs(raw$logFC) >= cut
  data.frame(lfc_cutoff = cut, n_up = sum(s & raw$logFC > 0), n_down = sum(s & raw$logFC < 0))
}))
print(sens, row.names = FALSE)
```

> **IMPROVEMENT #6 — `--padj` vs `--padj-cutoff` is a trap.**
> `--padj` is the **term-level** cutoff ([go_enrichment.R:71]); `--padj-cutoff`
> is the **gene-level** one ([:53]). Typing one when you meant the other is
> silent and changes the result. Rename to `--term-padj` / `--gene-padj`, or at
> minimum echo both into the console banner.

---

# 4. End-to-end CLI runs

```{r runner}
# `expect_msg` is a regex the output MUST match. Without it a test can pass for
# the wrong reason -- e.g. every run exits 1 when a package is missing, which
# would make the whole argument-validation section look green.
run_script <- function(label, args, expect_status = 0L, expect_msg = NULL) {
  t0  <- Sys.time()
  out <- suppressWarnings(system2("Rscript", c(shQuote(SCRIPT), args),
                                  stdout = TRUE, stderr = TRUE))
  st  <- attr(out, "status"); st <- if (is.null(st)) 0L else as.integer(st)
  txt <- paste(out, collapse = "\n")
  cat(sprintf("\n===== %s  (exit %d, %.1fs) =====\n", label, st,
              as.numeric(difftime(Sys.time(), t0, units = "secs"))))
  cat(txt, "\n")

  # An environment failure is not a test result.
  if (grepl("there is no package called|missing R packages", txt)) {
    skip(label, "package missing", part = "e2e")
    return(invisible(list(status = st, output = out, skipped = TRUE)))
  }
  ok <- st == expect_status &&
        (is.null(expect_msg) || grepl(expect_msg, txt, ignore.case = TRUE))
  chk(sprintf("'%s' -> exit %d%s", label, expect_status,
              if (is.null(expect_msg)) "" else sprintf(" + message /%s/", expect_msg)),
      ok, part = "e2e")
  invisible(list(status = st, output = out, skipped = FALSE))
}

expect_files <- function(label, dir, patterns) {
  if (!dir.exists(dir)) { skip(label, "no output dir", part = "e2e"); return(invisible(NULL)) }
  f <- list.files(dir)
  miss <- patterns[!vapply(patterns, function(p) any(grepl(p, f)), logical(1))]
  chk(sprintf("%s: expected outputs present", label), length(miss) == 0, part = "e2e")
  if (length(miss)) cat("  MISSING:", paste(miss, collapse = ", "), "\n")
  invisible(f)
}
```

`HAVE_OPTPARSE` gates the argument-validation block below: without optparse the
script aborts at `library(optparse)` and *every* run exits 1, which would make
these tests pass for entirely the wrong reason.

```{r optparse-gate}
HAVE_OPTPARSE <- requireNamespace("optparse", quietly = TRUE)
cat("optparse installed:", HAVE_OPTPARSE, "\n")
```

## 4.1 Argument validation (runs without Bioconductor)

These should all fail fast with exit status 1, before any package is loaded.

```{r e2e-validation, eval=HAVE_OPTPARSE}
run_script("no input at all", c("--outdir", file.path(OUTDIR, "v1")),
           1L, "supply either --genes")
run_script("--genes and --de-table together",
           c("--genes", file.path(FIXDIR, "up.txt"),
             "--de-table", deseq_csv, "--outdir", file.path(OUTDIR, "v2")),
           1L, "mutually exclusive")
run_script("--gsea without --de-table",
           c("--genes", file.path(FIXDIR, "up.txt"), "--gsea",
             "--background", file.path(FIXDIR, "background.txt"),
             "--outdir", file.path(OUTDIR, "v3")),
           1L, "requires --de-table")
run_script("invalid --ont",
           c("--genes", file.path(FIXDIR, "up.txt"), "--ont", "XX",
             "--outdir", file.path(OUTDIR, "v4")),
           1L, "--ont must be")
run_script("invalid --split",
           c("--genes", file.path(FIXDIR, "up.txt"), "--split", "sideways",
             "--outdir", file.path(OUTDIR, "v5")),
           1L, "--split must be")
run_script("missing --background is refused",
           c("--genes", file.path(FIXDIR, "up.txt"),
             "--outdir", file.path(OUTDIR, "v6")),
           1L, "no --background supplied")
run_script("--labels count mismatch",
           c("--genes", paste(file.path(FIXDIR, c("up.txt", "down.txt")), collapse = ","),
             "--labels", "only_one",
             "--background", file.path(FIXDIR, "background.txt"),
             "--outdir", file.path(OUTDIR, "v8")),
           1L, "--labels has")
```

```{r e2e-validation-skip, eval=!HAVE_OPTPARSE, echo=FALSE}
skip("argument validation suite", "optparse not installed", part = "e2e")
```

> **IMPROVEMENT #7 — validate `--p-method` and `--id-type` too.**
> `--ont` and `--split` are checked at [go_enrichment.R:133-136], but a typo in
> `--p-method` or `--id-type` surfaces much later as an opaque error from inside
> clusterProfiler. Add them to the same block:
>
> ```r
> if (!g("p-method") %in% p.adjust.methods) fail("--p-method must be one of: %s",
>     paste(p.adjust.methods, collapse = ", "))
> if (!g("id-type") %in% c("auto","SYMBOL","ENSEMBL","ENTREZID","REFSEQ","UNIPROT"))
>   fail("--id-type must be auto, SYMBOL, ENSEMBL, ENTREZID, REFSEQ or UNIPROT.")
> ```

```{r e2e-validation-improve, eval=HAVE_OPTPARSE}
run_script("IMPROVEMENT #7: bogus --p-method rejected early",
           c("--genes", file.path(FIXDIR, "up.txt"),
             "--background", file.path(FIXDIR, "background.txt"),
             "--p-method", "not-a-method",
             "--outdir", file.path(OUTDIR, "v7")),
           1L, "--p-method must be")
```

## 4.2 Plain gene list vs background

```{r e2e-list, eval=HAVE_BIOC}
d <- file.path(OUTDIR, "list_basic")
run_script("single list + background",
  c("--genes", file.path(FIXDIR, "up.txt"),
    "--background", file.path(FIXDIR, "background.txt"),
    "--organism", ORGANISM, "--ont", "BP",
    "--outdir", d, "--prefix", "UP"))
expect_files("single list", d,
             c("_all_terms\\.tsv$", "_significant\\.tsv$",
               "_dotplot\\.png$", "_dotplot\\.pdf$",
               "_summary\\.tsv$", "_run_log\\.txt$"))
```

### Validate the derived columns

```{r e2e-list-validate, eval=HAVE_BIOC}
f <- list.files(file.path(OUTDIR, "list_basic"), pattern = "_all_terms\\.tsv$", full.names = TRUE)
if (length(f)) {
  tt <- read.delim(f[1], check.names = FALSE)
  cat("terms:", nrow(tt), " columns:", paste(names(tt), collapse = ", "), "\n\n")

  chk("AnnotationSize/FoldEnrichment/Depth were added",
      all(c("AnnotationSize", "FoldEnrichment", "Depth") %in% names(tt)), part = "e2e")

  num <- as.numeric(sub("/.*", "", tt$GeneRatio)); den <- as.numeric(sub(".*/", "", tt$GeneRatio))
  bnum <- as.numeric(sub("/.*", "", tt$BgRatio));  bden <- as.numeric(sub(".*/", "", tt$BgRatio))
  chk("FoldEnrichment reproduces (k/n)/(M/N)",
      isTRUE(all.equal((num/den)/(bnum/bden), tt$FoldEnrichment)), part = "e2e")
  chk("AnnotationSize == numerator of BgRatio",
      isTRUE(all.equal(tt$AnnotationSize, bnum)), part = "e2e")
  chk("all terms respect --min-size/--max-size",
      all(tt$AnnotationSize >= 10 & tt$AnnotationSize <= 500), part = "e2e")
  chk("Depth is populated for BP", mean(!is.na(tt$Depth)) > 0.95, part = "e2e")
  chk("results are sorted by p.adjust", !is.unsorted(tt$p.adjust), part = "e2e")
  chk("raw p <= adjusted p everywhere", all(tt$pvalue <= tt$p.adjust + 1e-12), part = "e2e")

  cat("\nmedian annotation size:", median(tt$AnnotationSize, na.rm = TRUE),
      " median depth:", median(tt$Depth, na.rm = TRUE), "\n")
  head(tt[, c("ID","Description","GeneRatio","BgRatio","FoldEnrichment",
              "AnnotationSize","Depth","p.adjust")], 10)
}
```

> **IMPROVEMENT #8 — guard the column contract.**
> `annotate_terms` ([go_enrichment.R:317-330]) hard-codes `BgRatio` and
> `GeneRatio`. Recent DOSE releases already emit their own `FoldEnrichment` and
> `RichFactor` columns (which this code silently overwrites), and the ratio
> column names have shifted before. Add an explicit check so a version bump
> fails loudly instead of producing a wrong `AnnotationSize`:
>
> ```r
> req <- c("ID", "GeneRatio", "BgRatio", "Count", "pvalue", "p.adjust")
> if (!all(req %in% names(res_df)))
>   fail("enrichGO result is missing %s -- clusterProfiler %s may have renamed columns.",
>        paste(setdiff(req, names(res_df)), collapse = ", "),
>        as.character(packageVersion("clusterProfiler")))
> ```

## 4.3 `--simplify` and `--topgo`

```{r e2e-simplify, eval=HAVE_BIOC}
d <- file.path(OUTDIR, "list_simplify")
run_script("single list + --simplify",
  c("--genes", file.path(FIXDIR, "up.txt"),
    "--background", file.path(FIXDIR, "background.txt"),
    "--organism", ORGANISM, "--ont", "BP", "--simplify",
    "--outdir", d, "--prefix", "UP"))
expect_files("simplify", d, c("_significant_simplified\\.tsv$", "_dotplot_simplified\\.png$"))
```

```{r e2e-topgo, eval=HAVE_BIOC && HAVE_TOPGO}
d <- file.path(OUTDIR, "list_topgo")
run_script("single list + --topgo",
  c("--genes", file.path(FIXDIR, "up.txt"),
    "--background", file.path(FIXDIR, "background.txt"),
    "--organism", ORGANISM, "--ont", "BP", "--topgo",
    "--outdir", d, "--prefix", "UP"))

f <- list.files(d, pattern = "_all_terms\\.tsv$", full.names = TRUE)
if (length(f)) {
  tt <- read.delim(f[1], check.names = FALSE)
  chk("topGO columns merged in",
      all(c("topGO_weight01_p", "topGO_weight01_padj") %in% names(tt)), part = "e2e")
  chk("results still sorted after merge()", !is.unsorted(tt$p.adjust), part = "e2e")
  cat("terms significant in both ORA and topGO:",
      sum(tt$p.adjust < 0.05 & tt$topGO_weight01_padj < 0.05, na.rm = TRUE), "\n")
}
```

### EDIT #2 end-to-end

```{r e2e-topgo-bug, eval=HAVE_BIOC && HAVE_TOPGO}
r <- run_script("EDIT #2: --topgo with --allow-genome-background",
  c("--genes", file.path(FIXDIR, "up.txt"), "--allow-genome-background",
    "--organism", ORGANISM, "--ont", "BP", "--topgo",
    "--outdir", file.path(OUTDIR, "topgo_nobg"), "--prefix", "NOBG"), 0L)
```

## 4.4 Multi-list heatmap

```{r e2e-heatmap, eval=HAVE_BIOC}
d <- file.path(OUTDIR, "heatmap")
run_script("four lists + --heatmap",
  c("--genes", paste(file.path(FIXDIR, sprintf("list%d.txt", 1:4)), collapse = ","),
    "--labels", "L1,L2,L3,L4",
    "--background", file.path(FIXDIR, "background.txt"),
    "--organism", ORGANISM, "--ont", "BP", "--heatmap",
    "--outdir", d, "--prefix", "CMP"))
expect_files("heatmap", d,
             c("_term_clusters\\.tsv$", "_heatmap_matrix\\.tsv$",
               "_cross_list_heatmap\\.png$", "_cross_list_heatmap\\.pdf$"))

f <- list.files(d, pattern = "_term_clusters\\.tsv$", full.names = TRUE)
if (length(f)) {
  cl <- read.delim(f[1], check.names = FALSE)
  cat("terms:", nrow(cl), " clusters:", length(unique(cl$cluster)),
      " flagged representative:", sum(cl$is_representative), "\n")
  chk("EDIT #9: one representative per cluster",
      sum(cl$is_representative) == length(unique(cl$cluster)), part = "e2e")
  head(cl, 10)
}
```

## 4.5 DESeq2-shaped table, split up/down, with GSEA

```{r e2e-deseq, eval=HAVE_BIOC}
d <- file.path(OUTDIR, "deseq2")
run_script("DESeq2 table, --split both --gsea --rank-metric auto",
  c("--de-table", deseq_csv, "--gene-col", "gene_id",
    "--lfc-col", "log2FoldChange", "--padj-col", "padj", "--stat-col", "stat",
    "--organism", ORGANISM, "--ont", "BP", "--split", "both", "--gsea",
    "--outdir", d, "--prefix", "DE"))
expect_files("deseq2", d,
             c("DE_up_all_terms\\.tsv$", "DE_down_all_terms\\.tsv$",
               "DE_gsea_all_terms\\.tsv$", "DE_gsea_significant\\.tsv$",
               "DE_gsea_dotplot\\.png$", "DE_summary\\.tsv$"))
```

> **EDIT #11 — the README promises an automatic heatmap that never fires.**
> README:94 states "It also fires automatically for `--split both` on a DE
> table". [go_enrichment.R:534] gates it on `g("heatmap")` unconditionally. The
> run above produced no heatmap. Either fix the doc or change the gate to
> `if (g("heatmap") || (!is.null(g("de-table")) && g("split") == "both"))`.

```{r e2e-deseq-heatmap-bug, eval=HAVE_BIOC}
d <- file.path(OUTDIR, "deseq2")
hm <- list.files(d, pattern = "cross_list_heatmap")
cat("heatmap files produced by --split both without --heatmap:",
    if (length(hm)) paste(hm, collapse = ", ") else "(none)\n")
chk("EDIT #11: --split both produces a heatmap as README:94 claims",
    length(hm) > 0, part = "e2e")
```

> **EDIT #12 — the GSEA dotplot is PNG-only.**
> [go_enrichment.R:635] writes only `.png`; the ORA dotplot ([:362-363]) and the
> heatmap ([:458-459]) write both, and README:201 advertises `{png,pdf}`.
> Every figure in a manuscript needs the editable PDF.

```{r e2e-gsea-pdf-bug, eval=HAVE_BIOC}
chk("EDIT #12: GSEA dotplot PDF exists",
    length(list.files(file.path(OUTDIR, "deseq2"), pattern = "gsea_dotplot\\.pdf$")) > 0,
    part = "e2e")
```

## 4.6 edgeR table — both the broken and the fixed export

```{r e2e-edger-broken, eval=HAVE_BIOC}
r <- run_script("EDIT #1: raw topTags export (gene IDs in rownames)",
  c("--de-table", edger_raw,
    "--lfc-col", "logFC", "--padj-col", "FDR", "--stat-col", "F",
    "--organism", ORGANISM, "--ont", "BP", "--split", "both", "--gsea",
    "--outdir", file.path(OUTDIR, "edger_broken"), "--prefix", "BROKEN"), 0L)
cat("\nNote whether the message mentions 0 genes / an ID-mapping failure rather\n",
    "than the actual problem (an unnamed gene column).\n")
```

```{r e2e-edger-fixed, eval=HAVE_BIOC}
d <- file.path(OUTDIR, "edger_fixed")
run_script("edgeR table with an explicit gene column, --rank-metric auto",
  c("--de-table", edger_fix, "--gene-col", "gene_id",
    "--lfc-col", "logFC", "--padj-col", "FDR", "--stat-col", "F",
    "--organism", ORGANISM, "--ont", "BP", "--split", "both", "--gsea",
    "--outdir", d, "--prefix", "TE"))
```

Confirm `auto` chose `signed_sqrt` and that NES signs are not all one way:

```{r e2e-edger-check, eval=HAVE_BIOC}
f <- list.files(file.path(OUTDIR, "edger_fixed"), pattern = "gsea_all_terms\\.tsv$",
                full.names = TRUE)
if (length(f)) {
  gg <- read.delim(f[1], check.names = FALSE)
  cat("gene sets:", nrow(gg), "  NES > 0:", sum(gg$NES > 0), "  NES < 0:", sum(gg$NES < 0), "\n")
  chk("GSEA returns both positive and negative NES (sign was preserved)",
      sum(gg$NES > 0) > 0 && sum(gg$NES < 0) > 0, part = "e2e")
}
```

### EDIT #5 end-to-end

```{r e2e-edger-unsigned, eval=HAVE_BIOC}
d <- file.path(OUTDIR, "edger_unsigned")
run_script("EDIT #5: --rank-metric stat on the unsigned F column",
  c("--de-table", edger_fix, "--gene-col", "gene_id",
    "--lfc-col", "logFC", "--padj-col", "FDR", "--stat-col", "F",
    "--rank-metric", "stat",
    "--organism", ORGANISM, "--ont", "BP", "--split", "both", "--gsea",
    "--outdir", d, "--prefix", "UNSIGNED"))

f <- list.files(d, pattern = "gsea_all_terms\\.tsv$", full.names = TRUE)
if (length(f)) {
  gg <- read.delim(f[1], check.names = FALSE)
  cat("NES > 0:", sum(gg$NES > 0), "  NES < 0:", sum(gg$NES < 0),
      "  <- heavily one-sided means the sign was lost\n")
  chk("EDIT #5: script warned or corrected the unsigned metric",
      sum(gg$NES > 0) > 0 && sum(gg$NES < 0) > 0, part = "e2e")
}
```

## 4.7 EDIT #3 end-to-end — a DE table with no significant genes

```{r e2e-null, eval=HAVE_BIOC}
r <- run_script("EDIT #3: DE table with zero significant genes",
  c("--de-table", null_csv, "--gene-col", "gene_id",
    "--lfc-col", "log2FoldChange", "--padj-col", "padj", "--stat-col", "stat",
    "--organism", ORGANISM, "--ont", "BP", "--split", "both", "--gsea",
    "--outdir", file.path(OUTDIR, "null_de"), "--prefix", "NULL"), 0L)
cat("\nCheck the 'Detected gene ID type:' line above -- with an empty query it\n",
    "falls through to SYMBOL regardless of what the table actually contains.\n")
```

> **IMPROVEMENT #9 — sets skipped for being too small vanish from the summary.**
> The `next` at [go_enrichment.R:470] fires before `summary_rows[[n]]` is
> assigned, so "too few mapped genes" and "ran fine, found nothing" are
> indistinguishable in `*_summary.tsv`. Emit a row with `terms_tested = 0` and a
> `status` column instead.

## 4.8 Provenance log

```{r e2e-log, eval=HAVE_BIOC}
f <- list.files(file.path(OUTDIR, "deseq2"), pattern = "_run_log\\.txt$", full.names = TRUE)
if (length(f)) {
  lg <- readLines(f[1])
  chk("log records the command line", any(grepl("^command:", lg)), part = "e2e")
  chk("log records package versions", any(grepl("clusterProfiler", lg)), part = "e2e")
  chk("log records the resolved ID type", any(grepl("resolved ID type", lg)), part = "e2e")
  chk("log records background size", any(grepl("background size", lg)), part = "e2e")
  chk("log includes sessionInfo()", any(grepl("R version", lg)), part = "e2e")
  cat(head(lg, 30), sep = "\n")
}
```

> **IMPROVEMENT #10 — options left at their NULL default are absent from the log.**
> optparse omits `NULL`-valued options from `opt`, so the parameter dump at
> [go_enrichment.R:659-660] silently skips `--background`, `--stat-col`,
> `--gene-col` and friends when unset. For a file whose stated purpose is
> reproducibility (README:239-240), the log should record every option with its
> *effective* value, including "(unset)".

---

# 5. Summary of suggested edits

```{r summary-table, echo=FALSE}
edits <- read.csv(text = '
id,severity,location,issue
EDIT #1,CRITICAL,"go_enrichment.R:179-187 & 252-259","Raw topTags/DESeq2 exports have a blank first header field; gcol becomes the empty string, the name-membership validation still passes, de[[gcol]] is NULL, and the table silently collapses to 0 rows"
EDIT #2,HIGH,"go_enrichment.R:489","--topgo with --allow-genome-background passes NULL to run_topgo; zero-length allGenes factor crashes topGOdata"
EDIT #3,HIGH,"go_enrichment.R:285-288","ID type guessed from significant genes only; empty DE result silently yields SYMBOL and breaks GSEA mapping"
EDIT #4,MEDIUM,"go_enrichment.R:192","guess_id_type indexes with the pre-filter length, re-introducing NAs and diluting detection"
EDIT #5,MEDIUM,"go_enrichment.R:554 & 562-567","is_unsigned only consulted in the auto branch; --rank-metric stat on F/LR silently discards direction"
EDIT #6,LOW,"go_enrichment.R:494-495","Raw logical subsetting propagates NA p.adjust into an all-NA row"
EDIT #7,LOW,"go_enrichment.R:597-598","tapply+sort yields a 1-d array rather than a named numeric vector"
EDIT #8,LOW,"go_enrichment.R:351 & 442","factor(levels=) errors if two terms share a Description (reachable with --ont ALL)"
EDIT #9,LOW,"go_enrichment.R:424-426","is_representative set after truncation to --heatmap-top; most clusters end up with no representative"
EDIT #10,LOW,"go_enrichment.R:215-217","Mapping-rate line prints unique-Entrez count with a unique-source-ID percentage"
EDIT #11,DOC,"go_enrichment.R:534 vs README:94","README claims the heatmap fires automatically for --split both; the code requires --heatmap"
EDIT #12,DOC,"go_enrichment.R:635 vs README:201","GSEA dotplot writes PNG only; everything else writes PNG+PDF"
EDIT #13,DOC,"go_enrichment.R:8-12","Header comment calls clusterProfiler the only scriptable tool of the four, then offers topGO in the next bullet. README:19 has it right (DAVID and Enrichr are web-only)"
EDIT #14,DOC,"README:198","The phrase (every term tested) overstates it: enrichGO returns only terms with >=1 query gene and within min/maxGSSize"
EDIT #15,DOC,"README:27 vs README:247","Wijesooriya et al. author order differs between the in-text mention and the reference list"
', stringsAsFactors = FALSE)
knitr::kable(edits, row.names = FALSE)
```

```{r summary-improvements, echo=FALSE}
imps <- read.csv(text = '
id,location,suggestion
IMPROVEMENT #1,environments/,"Add a conda recipe; the script has never run because the stack was never installed"
IMPROVEMENT #2,README:45-46,"GOSemSim is an Imports of clusterProfiler, never optional; only topGO is conditional"
IMPROVEMENT #3,go_enrichment.R:174,"Header-detection regex misses Gene.ID / GeneSymbol / ensembl_gene_id"
IMPROVEMENT #4,go_enrichment.R:406,"Kappa universe is the term-gene union, not the input list as in Metascape; document it, since --kappa 0.3 is inherited from Metascape"
IMPROVEMENT #5,README:218-230,"--lfc-cutoff (default 1) and --padj-cutoff are absent from the defaults table but define what up/down mean"
IMPROVEMENT #6,go_enrichment.R:53 & 71,"--padj vs --padj-cutoff are silently confusable; rename to --term-padj / --gene-padj"
IMPROVEMENT #7,go_enrichment.R:133-136,"Validate --p-method and --id-type alongside --ont and --split"
IMPROVEMENT #8,go_enrichment.R:317-330,"Assert the enrichGO column contract so a DOSE version bump fails loudly"
IMPROVEMENT #9,go_enrichment.R:470,"Sets skipped for <5 genes get no summary row; emit one with a status column"
IMPROVEMENT #10,go_enrichment.R:659-660,"Options left at NULL never reach the run log, undermining its reproducibility purpose"
', stringsAsFactors = FALSE)
knitr::kable(imps, row.names = FALSE)
```

```{r final-report}
test_report()
```

---

# 6. Ready-to-apply patches

Not applied automatically — review, then paste into `go_enrichment.R`.

## EDIT #1 (critical)

Replace `read_table_auto` ([go_enrichment.R:179-187]):

```r
read_table_auto <- function(path) {
  if (!file.exists(path)) fail("file not found: %s", path)
  sep <- if (grepl("\\.csv$", path, ignore.case = TRUE)) "," else "\t"
  df  <- read.delim(path, sep = sep, header = TRUE, check.names = FALSE,
                    stringsAsFactors = FALSE)
  if (ncol(df) == 1) df <- read.delim(path, sep = ",", header = TRUE,
                                      check.names = FALSE, stringsAsFactors = FALSE)
  # DESeq2/edgeR tables written with row.names = TRUE leave the first header
  # field blank (or R renames it X / V1 / ...1). Promote it to a real column.
  if (!nzchar(trimws(names(df)[1])) || names(df)[1] %in% c("X", "V1", "...1")) {
    names(df)[1] <- "gene_id"
    msg("  note: first column of %s had no name; treating it as 'gene_id'.", basename(path))
  }
  df
}
```

and tighten the column check ([go_enrichment.R:252-255]):

```r
  gcol <- if (!is.null(g("gene-col"))) g("gene-col") else names(de)[1]
  if (!nzchar(trimws(gcol)) || is.null(de[[gcol]]))
    fail("gene column '%s' is empty or absent in --de-table (have: %s)",
         gcol, paste(names(de), collapse = ", "))
  for (cc in c(g("lfc-col"), g("padj-col")))
    if (!cc %in% names(de)) fail("column '%s' not in --de-table (have: %s)", cc,
                                 paste(names(de), collapse = ", "))
```

## EDIT #2

In the ORA loop, before `run_topgo` ([go_enrichment.R:487-491]):

```r
  if (g("topgo")) {
    bg_topgo <- if (is.null(bg_entrez)) keys(OrgDb, keytype = "ENTREZID") else bg_entrez
    onts <- if (g("ont") == "ALL") c("BP", "MF", "CC") else g("ont")
    tg <- do.call(rbind, lapply(onts, function(o) run_topgo(q, bg_topgo, o)))
    res <- merge(res, tg, by = "ID", all.x = TRUE)
  }
```

## EDIT #3 and #4

```r
guess_id_type <- function(x) {
  x <- head(x[!is.na(x) & nzchar(x)], 200)
  if (!length(x)) return("SYMBOL")
  if (mean(grepl("^ENS[A-Z]*[GT]?\\d{6,}", x)) > 0.5) return("ENSEMBL")
  if (mean(grepl("^\\d+$", x))                > 0.8) return("ENTREZID")
  if (mean(grepl("^N[MR]_|^X[MR]_", x))       > 0.5) return("REFSEQ")
  "SYMBOL"
}
```

and at [go_enrichment.R:285-288] prefer the background, which is always populated:

```r
if (id_type == "auto") {
  id_src  <- if (!is.null(bg_raw)) bg_raw else unlist(sets, use.names = FALSE)
  id_type <- guess_id_type(id_src)
  msg("Detected gene ID type: %s (from %s)", id_type,
      if (!is.null(bg_raw)) "background" else "query")
}
```

Note this must move *below* the `bg_raw` assignment.

## EDIT #5

After `is_unsigned` is computed ([go_enrichment.R:554]):

```r
  if (rmet == "stat" && is_unsigned) {
    msg("  WARNING: --stat-col '%s' is unsigned; using sign(logFC)*sqrt(%s) instead.",
        statcol, statcol)
    rmet <- "signed_sqrt"
  }
```

## EDIT #6, #7, #8, #9, #12

```r
# EDIT #6 -- go_enrichment.R:495
sig <- res[which(keep), , drop = FALSE]

# EDIT #7 -- go_enrichment.R:598
rk <- tapply(metric, ent, function(v) v[which.max(abs(v))])
rk <- setNames(as.numeric(rk), names(rk))
rk <- sort(rk, decreasing = TRUE)

# EDIT #8 -- go_enrichment.R:351 and :442
df$Description <- factor(df$Description,
                         levels = rev(unique(df$Description[order(df$FoldEnrichment)])))

# EDIT #9 -- go_enrichment.R:424-426
reps <- memb[!duplicated(memb$cluster), ]
memb$is_representative <- memb$ID %in% reps$ID          # flag ALL cluster reps
reps <- head(reps[order(reps$best_padj), ], g("heatmap-top"))
memb$in_heatmap <- memb$ID %in% reps$ID                 # separate column

# EDIT #12 -- after go_enrichment.R:636
ggsave(paste0(base, "_dotplot.pdf"), p, width = 9,
       height = max(3, 0.32 * nrow(d) + 1.4))
```

## EDIT #10

```r
  msg("  %-12s %d input -> %d mapped to %d Entrez IDs (%.1f%%)",
      label, length(ids), length(unique(m[[from]])), length(out),
      100 * length(unique(m[[from]])) / max(1, length(ids)))
```

---

# 7. Cleanup

```{r cleanup, eval=FALSE}
unlink(TESTROOT, recursive = TRUE)
```
